---
title: "mnpro_notes_LW"
format: html
editor: visual
---

## MNPRO Notes

------------------------------------------------------------------------

**Prevalence in each age sex/class**

## Introduction

This analyzes Chronic Wasting Disease (CWD) prevalence based on age/sex classes from simulation data from a model output for DPA 197 in minnesota . The analysis includes calculating prevalence rates for different age/sex classes, generating boxplots for each year, and combining them into a single multi-panel figure. A summary table of mean, median, and standard deviation of prevalence is also provided.

### What the Plots Represent

The plots are boxplots showing the distribution of CWD prevalence (proportion of infected individuals) for each age/sex class (e.g., AdultMale, YearlingFemale) across simulation iterations for a given year. The y-axis is log-scaled to handle low prevalence values. Boxplots display quartiles, with jittered points for individual non-zero prevalences and white circles for means. The combined figure allows comparison across years, highlighting trends like increasing prevalence or class-specific patterns over time.

## Data Loading and Preparation

This section loads the data and computes prevalence rates.

```{r}
# Load necessary libraries
library(dplyr)
library(readr)

# Load the CSV file
data <- read_csv("C:/Users/lrw0046/Downloads/cwdinfdyDPA197_seedinf_1.csv")
```

### This file contains simulation data with population and CWD counts for different deer classes per year and iteration.

here is the calculation for shown prevalence as CWD count divided by population for each class, handling zero denominators.This chunk adds new columns for prevalence rates using mutate. The ifelse ensures no division by zero, setting prevalence to 0 if population is zero. This prepares the data for filtering and plotting.

```{r}
# Calculate prevalence by class per row
data <- data %>%
  mutate(
    Prev_AdultMale = ifelse(AdultMale > 0, AdultMaleCWD / AdultMale, 0),
    Prev_YearlingMale = ifelse(YearlingMale > 0, YearlingMaleCWD / YearlingMale, 0),
    Prev_FawnMale = ifelse(FawnMale > 0, FawnMaleCWD / FawnMale, 0),
    Prev_AdultFemale = ifelse(AdultFemale > 0, AdultFemaleCWD / AdultFemale, 0),
    Prev_YearlingFemale = ifelse(YearlingFemale > 0, YearlingFemaleCWD / YearlingFemale, 0),
    Prev_FawnFemale = ifelse(FawnFemale > 0, FawnFemaleCWD / FawnFemale, 0)
  )
```

Filter out rows with zero total CWD to focus on meaningful data.We filter the data to include only rows where TotalCWD \> 0, as zero-CWD iterations don't contribute to prevalence analysis. This reduces noise in visualizations.

```{r}
# Exclude iterations with zero total CWD
filtered_data <- data %>% filter(TotalCWD > 0)
```

## Data Reshaping for Plotting

Reshape the data into long format for easier plotting with ggplot2. The data is moved to long format using pivot_longer, creating columns for 'Class' (e.g., AdultMale) and 'Prevalence'. clean the class names, set a factor order for consistent plotting, and replace zero prevalences with NA to avoid issues on log scales.

```{r}
# LONG format for plotting
filtered_long <- filtered_data %>%
  dplyr::select(Year, starts_with("Prev_")) %>%
  pivot_longer(
    cols = -Year,
    names_to = "Class",
    values_to = "Prevalence"
  ) %>%
  mutate(
    Class = gsub("Prev_", "", Class)
  )

# Set class order
class_order <- c("AdultMale", "YearlingMale", "FawnMale", "AdultFemale", "YearlingFemale", "FawnFemale")
filtered_long$Class <- factor(filtered_long$Class, levels = class_order)

# Replace zeros with NA for log scale plotting
filtered_long <- filtered_long %>%
  mutate(
    Prevalence_plot = ifelse(Prevalence > 0, Prevalence, NA_real_)
  )

```

Set up dynamic log breaks for the y-axis. This calculates dynamic breaks for the log-scaled y-axis based on the range of non-zero prevalence values. It ensures the axis labels are appropriate (e.g., 0.001%, 0.01%) and handles edge cases like all-zero data. Calculate anova p-values per year.

```{r}
# Dynamic log10 breaks
eps <- 1e-6
min_pos <- suppressWarnings(min(filtered_long$Prevalence_plot, na.rm = TRUE))
min_pos <- ifelse(is.finite(min_pos), min_pos, eps)
max_pos <- suppressWarnings(max(filtered_long$Prevalence_plot, na.rm = TRUE))
max_pos <- ifelse(is.finite(max_pos) && max_pos > 0, max_pos, 1)
lo_dec <- floor(log10(min_pos))
hi_dec <- ceiling(log10(max_pos))
log_breaks <- 10^(seq(lo_dec, hi_dec, by = 1))

# Get unique years
unique_years <- sort(unique(filtered_long$Year))

# List to store individual plots with ANOVA p-values
plot_list <- list()

# Loop over each year and create separate plot with ANOVA p-value
for (yr in unique_years) {
  
  # Filter data for this year
  year_data <- filtered_long %>% filter(Year == yr)
  
  # Means per class for this year (use positive values)
  summary_stats_year <- year_data %>%
    group_by(Class) %>%
    summarise(
      mean_val = mean(Prevalence[Prevalence > 0], na.rm = TRUE),
      .groups = "drop"
    )
  
  # Compute max_y for positioning
  max_y <- suppressWarnings(max(year_data$Prevalence_plot, na.rm = TRUE))
  if (!is.finite(max_y)) max_y <- 1
  
  # Compute per-year ANOVA p-value (using log-transformed positive prevalences)
  year_stats_data <- year_data %>%
    filter(Prevalence > 0) %>%
    mutate(log_prev = log10(Prevalence))
  if (nrow(year_stats_data) > 0 && length(unique(year_stats_data$Class)) > 1) {
    year_anova <- aov(log_prev ~ Class, data = year_stats_data)
    p_val <- summary(year_anova)[[1]][["Pr(>F)"]][1]
    p_label <- paste("ANOVA p =", format(p_val, digits = 3, scientific = TRUE))
  } else {
    p_label <- "ANOVA: Insufficient data"
  }
```

## Generate Individual Plots

Create a list of plots, one per year. This loop iterates over each unique year, filters the data, calculates class means, and generates a boxplot. Plots are stored in a list for later combination. No statistical annotations are added to keep is clean looking. Now includes anova stattistics for each year.

```{r}
  # Plot for this year with ANOVA annotation
  p <- ggplot(
    year_data,
    aes(x = Class, y = Prevalence_plot, fill = Class)
  ) +
    geom_boxplot(
      width = 0.5,
      outlier.shape = NA,
      alpha = 0.9,
      color = "black",
      linewidth = 0.3,
      na.rm = TRUE
    ) +
    geom_point(
      position = position_jitter(width = 0.2),
      alpha = 0.20, size = 0.45, shape = 16, color = "black",
      na.rm = TRUE
    ) +
    geom_point(
      data = summary_stats_year,
      aes(x = Class, y = mean_val, group = Class),
      size = 1.5, shape = 21, stroke = 0.35, fill = "white", color = "black",
      inherit.aes = FALSE,
      na.rm = TRUE
    ) +
    scale_fill_brewer(palette = "Set2") +
    scale_y_log10(
      breaks = log_breaks,
      labels = scales::percent_format(accuracy = 0.001),
      limits = c(max(eps, 10^lo_dec), max_y * 1.1)
    ) +
    labs(
      title = paste("Year", yr),
      x = "Age/Sex Class",
      y = "Prevalence"
    ) +
    annotate("text", x = Inf, y = Inf, label = p_label, hjust = 1.1, vjust = 1.1, size = 3, color = "red") +
    theme_minimal(base_size = 10) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.minor = element_blank(),
      legend.position = "none"
    )
  
  plot_list[[as.character(yr)]] <- p
}
```

Combine the plots into a single figure.

```{r}
# Combine all individual plots into one multi-panel figure
combined_plot <- grid.arrange(grobs = plot_list, ncol = 3, bottom = "Age/Sex Class", left = "Prevalence")

# Save the combined individual plots
ggsave("combined_cwd_prevalence_plots_with_stats.png", combined_plot, width = 15, height = 20, dpi = 300)

# Print the combined plot
print(combined_plot)

# Single plot with all years laid out together (x = Year, grouped boxplots by Class, no facets or stats annotation)
singular_plot_data <- filtered_long %>%
  mutate(Year = factor(Year))  # Treat Year as factor for grouping

singular_plot <- ggplot(
  singular_plot_data,
  aes(x = Year, y = Prevalence_plot, fill = Class)
) +
  geom_boxplot(
    position = position_dodge(width = 0.8),
    width = 0.7,
    outlier.shape = NA,
    alpha = 0.9,
    color = "black",
    linewidth = 0.3,
    na.rm = TRUE
  ) +
  geom_point(
    position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8),
    alpha = 0.20, size = 0.45, shape = 16, color = "black",
    na.rm = TRUE
  ) +
  # Add means (calculated per class and year)
  geom_point(
    data = filtered_long %>%
      filter(Prevalence > 0) %>%
      group_by(Year, Class) %>%
      summarise(mean_val = mean(Prevalence, na.rm = TRUE), .groups = "drop"),
    aes(x = factor(Year), y = mean_val, group = Class),
    position = position_dodge(width = 0.8),
    size = 1.5, shape = 21, stroke = 0.35, fill = "white", color = "black",
    na.rm = TRUE
  ) +
  scale_fill_brewer(palette = "Set2") +
  scale_y_log10(
    breaks = log_breaks,
    labels = scales::percent_format(accuracy = 0.001),
    limits = c(max(eps, 10^lo_dec), max_pos * 1.1)
  ) +
  labs(
    title = "CWD Prevalence by Year and Age/Sex Class",
    x = "Year",
    y = "Prevalence",
    fill = "Age/Sex Class"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )

# Save the singular plot
ggsave("singular_cwd_prevalence_plot.png", singular_plot, width = 15, height = 8, dpi = 300)

# Print the singular plot
print(singular_plot)
```

## Summary results

Generate a summary table of prevalence statistics.

```{r}
# Per-year ANOVA for Excel
stats_data <- filtered_long %>%
  filter(Prevalence > 0) %>%
  mutate(log_prev = log10(Prevalence))

per_year_anova <- stats_data %>%
  group_by(Year) %>%
  do(anova = tidy(aov(log_prev ~ Class, data = .))) %>%
  unnest(anova)

# Overall ANOVA tidy
overall_anova <- aov(log_prev ~ Class * Year, data = stats_data)
overall_anova_tidy <- tidy(overall_anova)

# Combine results for Excel
stats_list <- list(
  Overall_ANOVA = overall_anova_tidy,
  Per_Year_ANOVA = per_year_anova
)

# Write to Excel file
write_excel_csv(stats_list, "statistical_tests_results.csv")

# Print summaries (for console)
print("Overall ANOVA Summary:")
print(overall_anova_tidy)
print("Per-Year ANOVA Summaries:")
print(per_year_anova)
```

## Results Interpretation

results suggest CWD starts sporadically (non-significant tests, low effect sizes) but becomes structured over time, with higher prevalence in adults and males (significant tests, large d).

The summary table from the model outputs for DPA 197 reveals a clear progression of CWD prevalence across the 15-year simulation period, with distinct patterns by age and sex classes. Overall, prevalence escalates from minimal levels in Year 1 (mostly below 0.001 across classes) to substantially higher rates by Year 15 (ranging from 0.015 to 0.043), indicating a compounding effect as the disease establishes itself in the modeled population.

By age, adults exhibit the highest prevalence throughout, with AdultMale and AdultFemale means reaching 0.043 and 0.026 in Year 15, respectively, about 2–3 times those of fawns (FawnMale 0.015, FawnFemale 0.017). Yearlings fall in between, with means around 0.018 in later years. This difference (adults \> yearlings \> fawns) strengthens over time, as seen in the widening gaps: in Year 5, adult means are 2–4 times fawn means, but by Year 15, the ratio stabilizes at 2–3 times, showing age-related accumulation of infections.

Sex differences are evident but more pronounced in adults, where males show higher means than females (e.g., Year 15 AdultMale 0.043 vs. AdultFemale 0.026). In yearlings and fawns, sex patterns are less consistent, with near-equality in yearlings (both \~0.018 in Year 15) and occasional female edges in fawns (FawnFemale 0.017 vs. FawnMale 0.015). Standard deviations parallel means, rising from near-zero in early years to 0.01–0.027 in Year 15, with adults displaying the greatest dispersion (e.g., AdultMale SD 0.027).

------------------------------------------------------------------------

# Prevalence Insights

Code block written to determine out of intial outputs for model DPA 197, which iterations hit a prevalence of 1%, what year they hit 1% within those iterations, and for what year in those iterations that cwd was first detected.

```{r}
# Load the required libraries
library(dplyr)

# Read the data
data <- read.csv("cwdinfdyDPA197_seedinf_1.csv")

# Calculate true prevalence
data <- data %>%
  mutate(true_prevalence = TotalCWD / TotalPreHarvestPop)

# Find iterations that reach at least 1% prevalence at any point
iterations_to_keep <- data %>%
  group_by(iteration) %>%
  summarise(max_prev = max(true_prevalence, na.rm = TRUE)) %>%
  filter(max_prev >= 0.01) %>%
  pull(iteration)

# Filter the data to only those iterations
filtered_data <- data %>%
  filter(iteration %in% iterations_to_keep)

# For each iteration, find the first year where prevalence >= 1%
first_prevalence_years <- filtered_data %>%
  filter(true_prevalence >= 0.01) %>%
  group_by(iteration) %>%
  summarise(first_year_1pct = min(Year, na.rm = TRUE))

# For each iteration, find the first year where DetProb hits 1 (assuming DetProb >= 1 indicates detection)
first_detection_years <- filtered_data %>%
  filter(DetProb >= 1) %>%
  group_by(iteration) %>%
  summarise(first_year_detected = min(Year, na.rm = TRUE))

# Join the two data frames for the detection table
detection_results <- first_prevalence_years %>%
  left_join(first_detection_years, by = "iteration") %>%
  arrange(iteration)

# Display the detection results
cat("Detection Table:\n")
print(detection_results)


# Join back to get the rows for those first years for the average testing
first_year_rows <- filtered_data %>%
  inner_join(first_prevalence_years, by = c("iteration", "Year" = "first_year_1pct"))

# Calculate averages for each tested class and round to nearest whole number
averages <- first_year_rows %>%
  summarise(
    `Adult Male` = round(mean(AdultMaleTested, na.rm = TRUE)),
    `Yearling Male` = round(mean(YearlingMaleTested, na.rm = TRUE)),
    `Fawn Male` = round(mean(FawnMaleTested, na.rm = TRUE)),
    `Adult Female` = round(mean(AdultFemaleTested, na.rm = TRUE)),
    `Yearling Female` = round(mean(YearlingFemaleTested, na.rm = TRUE)),
    `Fawn Female` = round(mean(FawnFemaleTested, na.rm = TRUE))
  )

# To make it a table format, transpose it
averages_table <- data.frame(
  `Age/Sex Class` = names(averages),
  `Average CWD Tested` = as.vector(t(averages))
)

# Display the average testing table
cat("\nAverage Testing Table:\n")
print(averages_table)


```

```         
```

------------------------------------------------------------------------

### Comparison Outputs

1.  objective is to compare the deer that were sampled for CWD in leech lake and white earth to the model outputs.
    1.  Link to sample done in leech lake and white earth -\><https://github.com/anyadoc/WELLOvCWD/blob/main/Deer%20Harvest%20and%20Population%20Reports/CWD%20Surveillance%20Data.csv>

**Compare average from model data to data provided by Tiffany**

-   Model average is referred to as Model_average and data provided by Tiffany is referred to as Surveillance_average in the table.

-   Differences are provided both numerically and percentage wise between model data and surveillance data

```{r}
library(readr)
library(dplyr)
library(writexl)

# Read the model output CSV files
model197 <- read_csv("cwdinfdyDPA197_seedinf_1.csv")
model297 <- read_csv("cwdinfdyDPA297298_seedinf_1.csv")

# Compute average annual tested for each age/sex class in the models
model197_avg <- model197 %>%
  summarise(
    AdultMale = mean(AdultMaleTested, na.rm = TRUE),
    YearlingMale = mean(YearlingMaleTested, na.rm = TRUE),
    FawnMale = mean(FawnMaleTested, na.rm = TRUE),
    AdultFemale = mean(AdultFemaleTested, na.rm = TRUE),
    YearlingFemale = mean(YearlingFemaleTested, na.rm = TRUE),
    FawnFemale = mean(FawnFemaleTested, na.rm = TRUE)
  )

model297_avg <- model297 %>%
  summarise(
    AdultMale = mean(AdultMaleTested, na.rm = TRUE),
    YearlingMale = mean(YearlingMaleTested, na.rm = TRUE),
    FawnMale = mean(FawnMaleTested, na.rm = TRUE),
    AdultFemale = mean(AdultFemaleTested, na.rm = TRUE),
    YearlingFemale = mean(YearlingFemaleTested, na.rm = TRUE),
    FawnFemale = mean(FawnFemaleTested, na.rm = TRUE)
  )

# Hardcoded average annual surveillance data (total over 4 years / 4)
# For Leech Lake (DPA 197)
surv_avg197 <- c(
  AdultMale = 139 / 4,
  YearlingMale = 55 / 4,
  FawnMale = 83 / 4,
  AdultFemale = 229 / 4,
  YearlingFemale = 37 / 4,
  FawnFemale = 69 / 4
)

# For White Earth (DPA 297/298)
surv_avg297 <- c(
  AdultMale = 182 / 4,
  YearlingMale = 18 / 4,
  FawnMale = 27 / 4,
  AdultFemale = 120 / 4,
  YearlingFemale = 7 / 4,
  FawnFemale = 20 / 4
)

# Classes for dataframe
classes <- c("Adult Male", "Yearling Male", "Fawn Male", "Adult Female", "Yearling Female", "Fawn Female")

# Create dataframes for surveillance averages
surv_df197 <- data.frame(
  Class = classes,
  Surveillance_Avg = surv_avg197
)

surv_df297 <- data.frame(
  Class = classes,
  Surveillance_Avg = surv_avg297
)

# Create dataframes for model averages (transposed to match format)
model_df197 <- data.frame(
  Class = classes,
  Model_Avg = c(model197_avg$AdultMale, model197_avg$YearlingMale, model197_avg$FawnMale,
                model197_avg$AdultFemale, model197_avg$YearlingFemale, model197_avg$FawnFemale)
)

model_df297 <- data.frame(
  Class = classes,
  Model_Avg = c(model297_avg$AdultMale, model297_avg$YearlingMale, model297_avg$FawnMale,
                model297_avg$AdultFemale, model297_avg$YearlingFemale, model297_avg$FawnFemale)
)

# Compute comparisons for DPA 197
comparison197 <- data.frame(
  Class = classes,
  Model_Avg = model_df197$Model_Avg,
  Surveillance_Avg = surv_df197$Surveillance_Avg,
  Numerical_Difference = model_df197$Model_Avg - surv_df197$Surveillance_Avg,
  Percentage_Difference = ((model_df197$Model_Avg - surv_df197$Surveillance_Avg) / surv_df197$Surveillance_Avg) * 100
)

# Compute comparisons for DPA 297/298
comparison297 <- data.frame(
  Class = classes,
  Model_Avg = model_df297$Model_Avg,
  Surveillance_Avg = surv_df297$Surveillance_Avg,
  Numerical_Difference = model_df297$Model_Avg - surv_df297$Surveillance_Avg,
  Percentage_Difference = ((model_df297$Model_Avg - surv_df297$Surveillance_Avg) / surv_df297$Surveillance_Avg) * 100
)

# Write to Excel file
write_xlsx(
  list(
    "Model_Outputs_DPA197" = model_df197,
    "Model_Outputs_DPA297298" = model_df297,
    "MN_Surveillance_DPA197" = surv_df197,
    "MN_Surveillance_DPA297298" = surv_df297,
    "Comparisons_DPA197" = comparison197,
    "Comparisons_DPA297298" = comparison297
  ),
  "cwd_comparison_output.xlsx"
)
```

------------------------------------------------------------------------

Code Block to determine the number of rows within the intial model outputs that contain prevelances .5%, 1%, 2%, 3%, 4%, 5%, prevelances are within a .1 range of the specified prevelaces percantages. Prevealnces are true prevelances calculated by dividing total cwd / total preharvest pop.

```{r}
library(readr)
library(dplyr)
library(writexl)
library(ggplot2)
# Read the CSV files
dpa197 <- read_csv("cwdinfdyDPA197_seedinf_1.csv")
dpa297 <- read_csv("cwdinfdyDPA297298_seedinf_1.csv")
# Calculate true prevalence in percentage and add DPA label
dpa197 <- dpa197 %>%
  mutate(true_prev = (TotalCWD / TotalPreHarvestPop) * 100,
         DPA = "DPA197")
dpa297 <- dpa297 %>%
  mutate(true_prev = (TotalCWD / TotalPreHarvestPop) * 100,
         DPA = "DPA297")
# Combine the datasets
combined_df <- bind_rows(dpa197, dpa297)
# Define the target prevalences
targets <- c(0.5, 1.0, 2.0, 3.0, 4.0, 5.0)
# Function to assign target prevalence bin if within ±0.1%
assign_bin <- function(prev, targets) {
  for (t in targets) {
    if (prev >= (t - 0.1) & prev <= (t + 0.1)) {
      return(t)
    }
  }
  return(NA)
}
# Apply the bin assignment
combined_df <- combined_df %>%
  rowwise() %>%
  mutate(Target_Prevalence = assign_bin(true_prev, targets)) %>%
  ungroup() %>%
  filter(!is.na(Target_Prevalence)) # Filter to only rows in a bin
# Write the combined and filtered data to a CSV file
write_csv(combined_df, "combined_prevalence_data.csv")
# Create a static violin plot using ggplot2 with clean aesthetics, grouped by target
p <- ggplot(combined_df, aes(x = factor(Target_Prevalence), y = true_prev, fill = DPA)) +
  geom_violin(trim = FALSE, alpha = 0.6, linewidth = 0.5, position = position_dodge(0.9)) + # Dodged for grouping
  scale_fill_manual(values = c("DPA197" = "#1f77b4", "DPA297" = "#2ca02c")) +
  labs(title = "Violin Plots of True Prevalence Within ±0.1% of Each Target",
       x = "Target Prevalence (%)",
       y = "True Prevalence (%)",
       fill = "DPA Group") +
  theme_minimal(base_size = 14) + # Clean minimal theme
  theme(plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
        axis.title = element_text(size = 14),
        axis.text = element_text(size = 12),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 11),
        legend.position = "bottom", # Horizontal legend below
        panel.grid.major = element_line(color = "#e0e0e0"),
        panel.grid.minor = element_blank(),
        plot.margin = margin(10, 10, 20, 10)) +
  coord_cartesian(ylim = c(0, NA)) # Ensure y starts at 0, expandable upward
# Optionally overlay narrow box plots for quartiles (uncomment to enable)
# p <- p + geom_boxplot(width = 0.1, fill = "white", color = "black", outlier.shape = NA, position = position_dodge(0.9))
# Display the plot
print(p)
# To save as PNG (high resolution)
ggsave("prevalence_target_violin_static.png", p, width = 10, height = 6, dpi = 300)
```
